Prediction of Protein-Protein Interactions on the Human and Rice Interactome

### 2020-11-18 nicolas/PPIPL3/ReportV3.tex

Proteins are the key actors of biological processes inside cells.
As they carry out a variety of tasks, proteins do not work as single
agents but rather as a part of dynamic networks of protein-protein
interactions (PPI) \cite{Lin2017}. PPI networks play a key role in
a variety of interdependent mechanisms, including signal transduction,
homeostasis control, stress responses, plant defense and organ formation.
At the molecular level, PPIs play an essential role in many physiological
and developmental processes, including protein phosphorylation, transcriptional
co-factor recruitment and transporter activation \cite{Zhang_2010_PPI}.

Prediction of potentially relevant yet unexplored PPIs is a current
research topic on bioinformatics. Several authors have proposed different
methods for extrapolating information from the existing PPI networks.
Kovacs et al. (2019) introduces an approach which relies on counting
paths of length 3 (L3) to predict interactions among proteins for
a variety of model organisms. The proposed approach outperform previous
methods in different PPI networks of yeast (\emph{S. cerevisiae}),
Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D.
melanogaster}), fission yeast (\emph{S. pombe}) and mouse (\emph{M.
musculus}), as well as for the human interactome \cite{Kovacs2019}.

A common way to uncover (or validate) PPIs is the \emph{Yeast-Two-Hybrid}
technique (also known as \emph{two-hybrid screening} or \emph{Y2H}),
which is based on the expression of a specific reporter gene that
activates by the binding of a DNA-binding Domain (DB) and an Activation
Domain (AD) of a Transcription Factor that binds to an Upstream Activation
Sequence (UAS). For the Y2H technique, a protein is fused to the DB
domain (known as \emph{bait}) and another one to the AD (known as
\emph{prey}). If the proteins do not interact, then the reporter gene
is not expressed. Otherwise, the reporter gene expression is activated
by the activation domain.

@@

Based on the outcome of numerous Y2H experiments PPI networks can
be constructed, in which known interactions for each protein are represented.
Several algorithms are proposed over these networks in order to predict
yet unknown interactions. This paper evaluates predictions based on
three measures: Common Neighbors (\textbf{CN}), based on the count
of paths of length 2; the raw count of paths of length 3 (\textbf{A3});
and the degree-normalized count of paths of length 3 (\textbf{L3}).
Furthermore

@@

The focus of this paper is to evaluate different methods for predicting
PPIs using the existing knowledge of the network of interactions,
which is represented as an undirected graph. The traditional way of
approaching the problem is usually based on social networks analysis,
more specifically on the triadic closure principle (TCP), that states
that the more common shared friends that two people have, the more
likely that they know each other. As shown by previous studies, the
mentioned approach fails because it does not consider the structural
and chemical properties of the proteins \cite{Kovacs2019}.

For achieving the described results, human and rice PPI networks are
used and compared using state-of-the-art methods, as well as the proposed
ones (CN, A3, L3). In the case of the human network, human interactome
(\emph{HI-II-14}), as well as a curated version of it (\emph{HI-TESTED})
were used. A massive experimental assay was carried on and its results
were consolidated and used to build a validation network (\emph{HI-III}).

\begin{comment}
JF: read throughout the document but edited up here. I prefer to read
the results before editing the materials and methods section. 
\end{comment}

@@

Human interactome data and base source code were downloaded from the
repository of the length-3 degree normalized paths methodology \cite{Kovacs2019}:
the dataset \emph{HI-II-14} and \emph{HI-TESTED} are used for prediction
and the dataset \emph{HI-III} is used for validation.

Rice interactome information was downloaded from the STRING database
\cite{Szklarczyk2019}, corresponding to the \emph{Oryza sativa} subspecies.
The downloaded file was \emph{4530.protein.links.detailed.v11.0.txt}.
and contains more than 8 million PPIs from several resources. For
the purpose of this study and based on previous work, only PPIs with
evidence from curated databases were used (i.e. rows where the column
\emph{databases} has a value greater than zero), resulting in a network
with $5025$ nodes and $164420$ edges.

@@

Previous code implementation was adapted from \texttt{C++} to \texttt{Python}
(V3.6), in order to unify the algorithms into one single script. For
the purpose of algorithmic validation, the three methods were implemented
from scratch with basic functionalities and data structures of the
\texttt{Python} language.

@@

Next experiment carried out was to add each prediction method score
as a prediction feature, that is, use the score as the eleventh input
feature for the ML algorithm. Results for the Common Neighbors method
(\textbf{CN}), which uses paths of length 2, are shown in Figure \ref{F2}.
It can be observed again that all 10 experiments have small variability
among them and perform significantly better than the baseline, with
an area under the ROC curve of $0.90$, with small standard deviation.
When looking at the confusion matrix for this model, turns out that
most of the guesses are true positives and true negatives, resulting
in a precision of $0.9285$ and a recall of $0.7420$. Same analyses
are done with the count of paths of length 3 (\textbf{A3}) and with
the degree-normalized length-3 score (\textbf{L3}) and results are
presented in figures \ref{F3} and \ref{F4}, respectively, resulting
also in area under the curve of 0.9 for both cases.

@@

The precision when using the count of paths of length 3 (A3) is $0.8587$
while the recall is $0.7685$. It is interesting to observe that precision
decreased $7.5\%$ when compared to the previous case, but recall
increased $3.6\%$. A similar situation is obtained for the model
with the normalized score (L3), whose precision decreased to $0.8936$
($-3.8\%$) and recall increased to $0.7610$ ($+2.6\%$). It can
be seen also that the ROC curve rapidly achieves higher values of
true positive rate in the A2-featured model when compared to the other
models.

@@

Taking into account the different results validated in this report,
one can conclude that length-3 path methodologies might work better
on protein-protein interactions than its traditional length-2 (TCP
based) counterparts. On the other hand, it can be seen that degree-normalization
has little effect on the predictions, i.e., non-normalized $A{{}^3}$
matrix predictions are still a good methodology for edge prediction
on PPI networks.

>>>>

Proteins are key actors of biological processes inside cells.
They function as are part of dynamic networks of protein-protein
interactions (PPI) rather than carrying out a variety of tasks as isolated
agents~\cite{Lin2017}. At the cellular level, PPI networks play a key role in
a number of interdependent mechanisms, including signal transduction,
homeostasis control, stress responses, plant defense and organ formation.
At the molecular level, PPI networks also play an essential role in physiological
and developmental processes, including protein phosphorylation, transcriptional
co-factor recruitment and transporter activation~\cite{Zhang_2010_PPI}.

%Prediction of potentially relevant yet unexplored PPIs is a current research topic on bioinformatics. 
Several authors have introduced different
methods for extrapolating information from PPI networks.
Kovacs et al. (2019) proposes an approach which relies on counting
paths of length 3 (L3) to predict interactions among proteins for
a variety of model organisms. The proposed approach outperforms previous
methods for PPI networks of yeast (\emph{S. cerevisiae}),
Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D.
melanogaster}), fission yeast (\emph{S. pombe}) and mouse (\emph{M.
musculus}), as well as for the human interactome \cite{Kovacs2019}.

A common way to uncover (or validate) PPIs is the \emph{yeast-two-hybrid} or \emph{Y2H}
technique (also known as \emph{two-hybrid screening}). Y2H is based on the expression of a so called reporter gene that is
activated by the binding of a DNA-binding domain and an activation
domain of a transcription factor. This transcription factor then binds to an upstream activation
sequence. For the Y2H technique, a protein is fused to the DNA-binding domain (known as \emph{bait}) and another protein to the activation domain (known as
\emph{prey}). Only when the proteins interact, the reporter gene
is expressed. Otherwise, the reporter gene expression is activated
by the activation domain.

@@

PPI networks are constructed based on the outcome of numerous Y2H experiments in which known interactions for each protein are represented. 
Several algorithms are proposed over these networks in order to predict
hidden interactions. This paper evaluates PPI predictions based on
three measures: the count of paths of length 2 or Common Neighbors (\textbf{CN}); the raw count of paths of length 3 (\textbf{A3}); and the degree-normalized count of paths of length 3 (\textbf{L3}).

%Furthermore

@@

Our focus is to evaluate different methods for predicting
PPIs using the existing knowledge of the network of interactions,
which is represented as an undirected graph. 
\begin{comment}
    not sure we can call the TCP approach the traditional approach
\end{comment}

% The traditional way of approaching the problem is usually based on social networks analysis,
% more specifically on the triadic closure principle (TCP), that states
% that the more common shared friends that two people have, the more
% likely that they know each other. As shown by previous studies, the
% mentioned approach fails because it does not consider the structural
% and chemical properties of the proteins \cite{Kovacs2019}.

Human and rice PPI networks are used to compare state-of-the-art methods
%what are these state of the art methods?
, as well as the proposed approach (CN, A3, L3). In the case of the human network, the human interactome dataset denoted by \emph{HI-II-14} and its curated version (\emph{HI-TESTED})
were used for evaluation. Results on experimental assays are consolidated to build a validation network denoted by \emph{HI-III}.

\begin{comment}
JF: need to comment on rice as well  (in the same section). Or move the discussion above to Material and Methods

JF: read throughout the document but edited up here. I prefer to read
the results before editing the materials and methods section. 
\end{comment}

@@

Human interactome data and base source code were downloaded from the
repository of the length-3 degree normalized paths methodology \cite{Kovacs2019}:
the datasets \emph{HI-II-14} and \emph{HI-TESTED} are used for prediction
and the dataset \emph{HI-III} is used for validation.

\begin{comment}
    JF: I thought you were generating \emph{HI-III} based on your results - this is confusing

    \end{comment}

Rice interactome information was downloaded from the STRING database
\cite{Szklarczyk2019}, corresponding to the \emph{Oryza sativa} subspecies.
The file labelled \emph{4530.protein.links.detailed.v11.0.txt}.
contains more than 8 million PPIs from several sources. For
the purpose of this study, only PPIs with
evidence from curated databases are considered (i.e. rows where the column
\emph{databases} has a value greater than zero). The resulting network consists of
$5025$ nodes and $164420$ edges.

@@

Previous code implementation was adapted from \texttt{C++} to \texttt{Python}
(V3.6), in order to unify the algorithms into one single script. 
% from the perspective of the reader what other code was in Python?
For
the purpose of algorithmic validation, the three methods were implemented
from scratch with basic functionalities and data structures of the
\texttt{Python} language.

@@

Next experiment carried out was to add each prediction method score
as a prediction feature, that is, use the score as the eleventh input
feature for the ML algorithm. Results for the Common Neighbors method
(\textbf{CN}), which uses paths of length 2, are shown in Figure \ref{F2}.
It can be observed again that all 10 experiments have small variability
among them and perform significantly better than the baseline, with
an area under the ROC curve of $0.90$, with small standard deviation.
When looking at the confusion matrix for this model, turns out that
most of the guesses are true positives and true negatives, resulting
in a precision of $0.93$ and a recall of $0.74$. Same analyses
are done with the count of paths of length 3 (\textbf{A3}) and with
the degree-normalized length-3 score (\textbf{L3}) and results are
presented in figures \ref{F3} and \ref{F4}, respectively, resulting
also in area under the curve of 0.9 for both cases.

@@

The precision when using the count of paths of length 3 (A3) is $0.86$
while the recall is $0.77$. It is interesting to observe that precision
decreased $7.5\%$ when compared to the previous case, but recall
increased $3.6\%$. A similar situation is obtained for the model
with the normalized score (L3), whose precision decreased to $0.89$
($-3.8\%$) and recall increased to $0.76$ ($+2.6\%$). It can
be seen also that the ROC curve rapidly achieves higher values of
true positive rate in the A2-featured model when compared to the other
models.

@@

Taking into account the different results validated in this paper,
one can conclude that length-3 path methodologies 
% what are the L3 methodologies, in the introduction only one is mentioned
might work better
on protein-protein interactions than its traditional length-2 (TCP
based) counterparts. On the other hand, it can be seen that degree-normalization
has little effect on the predictions, i.e., non-normalized $A{{}^3}$
matrix predictions are still a good methodology for edge prediction
on PPI networks.
